---
title: "Assignment 10: Data Scraping"
author: "Haochuan Zhan"
output:
  pdf_document:
    latex_engine: xelatex
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(rvest)

getwd()
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"
the_website <- read_html(url)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
the_water_system_name_tag <- "div+ table tr:nth-child(1) td:nth-child(2)" 
the_contact_person_tag <- "div+ table tr:nth-child(4) td:nth-child(2)" 
the_ownership_tag <- "div+ table tr:nth-child(2) td:nth-child(4)"
the_max_day_use_tag <- "th~ td+ td"

water_system_name <- the_website %>% html_nodes(the_water_system_name_tag) %>% html_text(trim = TRUE) %>% .[1]
contact_person <- the_website %>% html_nodes(the_contact_person_tag) %>% html_text(trim = TRUE) %>% .[1]
ownership <- the_website %>% html_nodes(the_ownership_tag) %>% html_text(trim = TRUE) %>% .[1]
max_day_use <- the_website %>% html_nodes(the_max_day_use_tag) %>% html_text(trim = TRUE)

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 
df_withdrawals <- data.frame(
  Month = c("Jan","May","Sept","Feb","Jun","Oct","Mar","Jul","Nov","Apr","Aug","Dec"),
  Year  = rep(2024, 12),
  WaterSystemName = rep(water_system_name, 12),
  ContactPerson   = rep(contact_person, 12),
  Ownership       = rep(ownership, 12),
  MaxDayUse_MGD   = as.numeric(max_day_use),
  stringsAsFactors = FALSE
)

df_withdrawals$Date <- my(paste(df_withdrawals$Month, "-", df_withdrawals$Year))

df_withdrawals <- df_withdrawals[order(df_withdrawals$Date), ]

View(df_withdrawals)

#5 
plot(df_withdrawals$Date, df_withdrawals$MaxDayUse_MGD, type = "l",
     xlab = "Month (2024)", ylab = "Maximum Day Use (MGD)")

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output

```{r construct.a.scraping.function}
#6. 
scrape_lwsp <- function(PWSID, the_year){
  the_scrape_url <- paste0(
    "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=",
    PWSID,
    "&year=",
    the_year
  )
  the_website <- read_html(the_scrape_url)
  the_water_system_name_tag <- "div+ table tr:nth-child(1) td:nth-child(2)"
  the_contact_person_tag    <- "div+ table tr:nth-child(4) td:nth-child(2)"
  the_ownership_tag         <- "div+ table tr:nth-child(2) td:nth-child(4)"
  the_data_tag              <- "th~ td+ td"
  water_system_name <- the_website %>% html_nodes(the_water_system_name_tag) %>% html_text(trim = TRUE) %>% .[1]
  contact_person    <- the_website %>% html_nodes(the_contact_person_tag)    %>% html_text(trim = TRUE) %>% .[1]
  ownership         <- the_website %>% html_nodes(the_ownership_tag)         %>% html_text(trim = TRUE) %>% .[1]
  max_day_use       <- the_website %>% html_nodes(the_data_tag)              %>% html_text(trim = TRUE)
  df_withdrawals <- data.frame(
    Month = c("Jan","May","Sept","Feb","Jun","Oct","Mar","Jul","Nov","Apr","Aug","Dec"),
    Year  = rep(the_year, 12),
    PWSID = rep(PWSID, 12),
    MaxDayUse_MGD = as.numeric(max_day_use[1:12])
  )
  df_withdrawals <- df_withdrawals %>%
    mutate(
      WaterSystemName = !!water_system_name,
      ContactPerson   = !!contact_person,
      Ownership       = !!ownership,
      Date = my(paste(Month, "-", Year))
    )

  df_withdrawals <- df_withdrawals[order(df_withdrawals$Date), ]

  return(df_withdrawals)}
df1 <- scrape_lwsp("03-32-010", 2023)
View(df1)
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 
df_2020 <- scrape_lwsp("03-32-010", 2020)

plot(df_2020$Date, df_2020$MaxDayUse_MGD, type = "l",
     xlab = "Month (2020)", ylab = "Maximum Day Use (MGD)")

str(df_2020)
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 
# Cary 2020
df_cary_2020 <- scrape_lwsp("03-92-020", 2020)

# Combine with Durham 2020 (df_2020 from Q7)
df_both_2020 <- rbind(df_2020, df_cary_2020)
plot(df_2020$Date, df_2020$MaxDayUse_MGD, type = "l",
     ylim = c(15, max(c(df_2020$MaxDayUse_MGD, df_cary_2020$MaxDayUse_MGD), na.rm = TRUE) + 2),
     xlab = "Month (2020)", ylab = "Maximum Day Use (MGD)")
lines(df_cary_2020$Date, df_cary_2020$MaxDayUse_MGD)
legend("topleft", legend = c("Durham", "Cary"), lty = 1)
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 
years <- 2018:2023

df_cary_all <- map2_dfr(rep("03-92-020", length(years)), years, scrape_lwsp)

ggplot(df_cary_all, aes(x = Date, y = MaxDayUse_MGD)) +
  geom_line() +
  geom_smooth(method = "loess") +
  labs(x = "Month", y = "Maximum Day Use (MGD)")
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:Yes, it looks like Cary’s max daily water use shows a slight upward trend from 2018 to 2023. The monthly values bounce around a lot, but the smoothed line gradually increases over time and then levels off a bit near the end.
>
